---
title: "Masters Homework 2"
author: "Daniel Juarez"
date: "7/21/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Credit received ##

>- Find summary information about the total blood mercury levels in our NYC HANES data (hint: names(nyc) will call the names of each column). What is the max value, what is the minimum value.  

```{r}
NYC <- read.csv("NYC_HANES_DIAB.csv")
#names(NYC)  used for checking which name the column for mercury levels in blood
summary(NYC$MERCURYTOTALBLOOD)
minimum_value <- min(NYC$MERCURYTOTALBLOOD, na.rm = TRUE)
max_value <- max(NYC$MERCURYTOTALBLOOD, na.rm = TRUE)
print(minimum_value)
print(max_value)
```


>- Bonus/Challenge: What is the Standard Deviation and skewness of the blood mercury?

```{r}
blood_mercury <- na.omit(NYC$MERCURYTOTALBLOOD) #clean data
standard_dev <- sd(blood_mercury) #standard deviation
mean_blood_mercury <- mean(blood_mercury) #mean
length_blood_mercury <- length(blood_mercury) #length
skewness <- sum((blood_mercury - mean_blood_mercury)^3) / (length_blood_mercury * standard_dev^3) #put the variables together into the skewness calculation
print(standard_dev)
print(skewness)
```



>- Install a package and load, such as "ggplot2" or "dplyr"

```{r}
install.packages("ggplot2")#installs the package
library(ggplot2) #runs the program

```





>- What two commands are needed to install and use a package in R?

```{r}
#install.packages('name of package')
#to turn on your package you would use the following command: library(name of package)
```




>- When you read in a .csv file into R, what class is the object that you read in (like nyc)?

The class of of NYC.csv file would be a data frame because it is most similar to an excel sheet in that you have columns and rows with characters, numeric, and integers.

>- What command can you use to determine what type of data is in a column?

```{r}
class(NYC$MERCURYTOTALBLOOD) #use the class(.csvfilename$nameofcolumn) in this example We used the column MERCURYTOTALBLOOD
```


>- Provide a command to find the type of data in the DX_DBTS column of nyc. 

```{r}
class(NYC$DX_DBTS)
```


>- Provide the same command the type of data but use the other way of indexing from what you did above. 

```{r}
class(NYC[,c('DX_DBTS')])
```



>- What would happen if we had chosen the first 4 columns of NYC HANES and made a matrix? What would the class of the cells in this matrix be? Why?

```{r}
matrix_of_first_4_columns <- matrix(NYC[,c(1:4)])
class(matrix_of_first_4_columns)
#as shown when I create a matrix of the first four columns of NYC HANES the class of the cells are matrix and array. This is because now that the four columns are in a matrix, this is now classified as a matrix. It is no longer just one column and therefore not one type of data. It also puts out an array because a matrix is a 2d form of an array because it has rows and colmns.

```


>- If we make a vector,vec<- c(3,9,4,7,8,9), and multiply it by WannaMatrix, what is the value in the third row, third column? 

```{r}
WannaMatrix <- nyc[5:10, 6:9]
WannaMatrix <- as.matrix(WannaMatrix,nrow = 6, ncol = 4)
vec<- c(3,9,4,7,8,9)
matrixxvec <- WannaMatrix*vec
value <- matrixxvec[3,3]
print (value)
#print(matrixxvec) to view if correct


```



>- What happens if instead we made a vector vec2 <- c(3,9,4,7,8) and multiply it by WannaMatrix, what is the third row, third column value? What is going on? 

```{r}
vec2 <- c(3,9,4,7,8)
matrixxvec2 <- WannaMatrix*vec2
value2 <- matrixxvec[3,3]
print (value2)
#a warning occurs stating that the value2 new matrix is not a multiple of shorter object length. This is because you cannot multiply a vector by a matrix that are different lengths.
```


>- Look up the arguments for read.csv

```{r}
?read.csv
```


>- Use apply() to find the median value of CADMIUM, LEAD CHOLESTEROLTOTAL, GLUCOSESI, CREATININESI, and CREATININE

```{r}
cleanNYC <- na.omit(NYC)
apply((cleanNYC[,c('CADMIUM', 'LEAD', 'CHOLESTEROLTOTAL', 'GLUCOSESI', 'CREATININESI', 'CREATININE')]),2,median)
```


>- Do the same using sapply()

```{r}
sapply((cleanNYC[,c('CADMIUM', 'LEAD', 'CHOLESTEROLTOTAL', 'GLUCOSESI', 'CREATININESI', 'CREATININE')]),median)
```

>- Use a for loop to find the median value of columns 16:19

```{r}
for(i in 16:19){
  median_values <- median(nyc[,i], na.rm = TRUE)
  print(c(colnames(nyc[i]),median_values))
}

```


>- Make a for loop to find the standard deviation (hint sd() ) of CHOLESTEROLTOTAL, GLUCOSESI, CREATININESI, AND CREATININE

```{r}
for(i in 16:19){
  sd_values <- sd(nyc[,i], na.rm = TRUE)
  print(c(colnames(nyc[i]),sd_values))
}
```


>- Bonus/Challenge Create a vector in which to put the standard deviations 

```{r}
#empty vector:
std_vector <- c()
#for loop used in prior example, but to get the std values which are then input into the vector
for(i in 16:19){
  std_values <- sd(nyc[,i], na.rm = TRUE)
  std_vector<-c(std_vector,std_values)
}
print(std_vector)
is.vector(std_vector)
```


>- Use ifelse to create a new variable of high and normal A1C dichotomizing with 7 as a threshold for high levels

```{r}
nyc_1 <- na.omit(nyc)
nyc$A1C_levels <-ifelse(nyc$A1C  >= 7, "High", "Normal")
```


>- Subset to only have females (gender == "female")

```{r}

```


>- Create a subset of males and find the median total cholesterol level

```{r}

```


>- Try to create a summary of the mean A1C using two factors, GENDER and DX_DBTS

```{r}

```

Find the class of the first three columns of data. What are they?

```{r}

```


